---
title: "Statistical_Inference_Course_Project_Part_2"
author: Jason B.
output:
  html_document: default
  pdf_document: default
---

## Overview  
#### The purpose of this exercise is to perform basic inferential statistical analysis of the ToothGrowth data.    
####The documention states, "**The Effect of Vitamin C on Tooth Growth in Guinea Pigs: ** The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice or ascorbic acid (a form of vitamin C and coded as VC)."  

##Load the data
```{r}
library(ggplot2)
library(datasets)
data("ToothGrowth")
```

##Explore the data  
```{r}
str(ToothGrowth)
sum(is.na(ToothGrowth))
summary(ToothGrowth)
```
  
####There are 3 variables, length, supplement, and dose. There are no missing values and the data frame appears "tidy". The variable len and dose are numeric doubles and the variable supp is a factor with two levels.  

##Plot the data 
```{r}
ggplot(ToothGrowth, aes(dose, len))+
         geom_smooth(method = "lm")+
         geom_point()+
         facet_grid(.~supp)
```
  
####There is a general upward trend for tooth length as a function of dose. The slope of the regression line is steeper for delivery via vitamin C.  

```{r}
ggplot(ToothGrowth, aes(len))+
        geom_density()+
        facet_grid(dose~.)+
        ggtitle("PDF of tooth length split among doses of Vitamin C")
```
  
####These density plots show the distribution of tooth length values as a function of dose irrespective of delivery method. The data are somewhat normally distributed. I will assume a normal distribution for inferential analysis.  
  
##Inferential analysis
```{r}
t.test(len ~ supp, data = ToothGrowth, var.equal = F, alternative = "two.sided")
```
  
####This analysis compares the effect of delivery methods on tooth length. It is not clear if the authors had a directional hypothesis; so, I performed a two-tailed test. Also, to be conservative, I assumed unequal variances. The p-value is greater than alpha = 0.05; therefore, this analysis fails to reject the null hypothesis that there is no effect of delivery method.  
  
```{r}
#subset by dose
tg_low <- subset(ToothGrowth$len, ToothGrowth$dose == 0.5)
tg_mid <- subset(ToothGrowth$len, ToothGrowth$dose == 1.0)
tg_hi <- subset(ToothGrowth$len, ToothGrowth$dose == 2.0)
plm <- t.test(tg_low, tg_mid)$p.value
plh <- t.test(tg_low, tg_hi)$p.value
pmh <- t.test(tg_hi, tg_mid)$p.value
uncorr_p <- c(p_low_mid = plm, p_low_hi = plh, p_mid_hi = pmh)
corr_p <- 3*uncorr_p
corr_p
```
  
####This analysis comparing the effect of different doses of Vitamin C on tooth length reveals something interesting. The p-values from the comparisons of the low to mid, low to hi, and mid to hi doses are all less than alpha = 0.5. Again, the tests were performed assuming unequal variances and with a two-sided hypothesis. Note, that I used a Bonferroni correction to control for multiple hypothesis testing. Thus, this analysis rejects the null hypothesis. The average tooth length was `r mean(tg_low)` for the 0.5 dose, `r mean(tg_mid)` for the 1.0 dose, and `r mean(tg_hi)` for the 2.0 dose. Thus, higher doses were associated with increased length.
  
##Conclusion
#### In this study, my analysis suggests that increasing doses of Vitamin C increased the length of guinea pig teeth irrespective of delivery method. 